Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-04-02 paper/response-bioinformatics.txt

We agree with the reviewers that an application of our work in a
bioinformatics analysis would be highly beneficial. For example, once
sufficient data is annotated with the CPO, we can use its taxonomic
structure for enrichment analyses. Similarly, once more cellular
phenotypes are being annotated in various model organism databases, we
can utilize CPO for integrative phenotype analyses across
species. However, at the moment, no comprehensive ontology of
physiology phenotypes is available, partly due to the complexity of
the domain and partly due to limitations in available ontology
frameworks. We see our main contribution in the development of a
theory that allows to capture physiology phenotypes; we applied this
theory to construct (largely automatically) the CPO, which is now
being used to annotate cellular phenotypes in the systems microscopy
domain. Based on our contributions, it is now possible to annotate
and eventually integrate (through the use of a controlled vocabulary
of physiology phenotypes, of which the CPO provides a first example)
data in the domain of physiology. At the same time, we maintain
compatibility with GO (on which a majority of physiology phenotypes
are based) and other PATO-based phenotype ontologies, thereby filling
a gap that has limited data integration in the physiology domain so
far.

Reviewer 2:
A theory of physiology related phenotypes and an implementation based
on GO, GO cross-products and PATO is presented. An outline of a theory
of phenotypes is introduced which participate-in a set of pathological
processes which can be qualified by a small set of PATO qualities. At
least, for a majority of heterogeneous phenotype descriptions this can
be an approach for unification. In future work a sound formal
characterization has and an evaluation of the practical value of the
approach has to be delivered.

Response:
We have now added a detailed formal presentation of our axioms as
supplementary material. An evaluation of the practical utility depends
on data that is being annotated with physiology phenotypes. Currently,
no comprehensive ontology of physiology phenotypes is available. Our
work brings the immediate benefit that the availability of CPO enables
annotation of cellular phenotypes in the systems microscropy
community, and our method can be applied to physiology phenotypes in
other domains, in particular computational modelling.

Reviewer 2:
Minor revision (page 2 line 27): You introduce a Heart beating example
to illustrate the challenges in modeling processes and their
regulation. [...] The Heart example is illustrative but limited with
respect to your ontology. There, both the the function of the single
entity can be impaired and the regulation of multiple occurences of
processes (where the example ends due to the single heartedness of
many animals).

Response: 
In response to this comment and reviewer 1's comment, we have removed
the Heart beating example from the introduction and focus on the
detailed presentation of the processes involved in the Results section
of our manuscript.

Reviewer 2:
Minor revision (page 3 second column line 33) see as above on the
phases of a single Heart action. The heart has no phase of
“inactivity” (that would be death), only a number of phases of which
the filling/ ejection phases are the least granular.

Response: We have changed our manuscript

Reviewer 2:
Minor revision (page 5 Table 1): make it clear that you refer to the
product of row and column quantitites in the cells. I find it
difficult to understand this easy relation from your legend. Do you
really need this table?

Response:
We have removed the table from our manuscript, and discuss the
content of the table now in the main text.

Reviewer 2:
Class names should be consistently written italic and uppercase
throughout the manuscript (e.g. Heart beating and Rate of heart
beating)

Response:
We have made this change throughout our manuscript.

### 2012-04-04 trunk/paper/response-bioinformatics.txt

We agree with the reviewers that an application of our work in a
bioinformatics analysis would be highly beneficial. For example, once
sufficient data is annotated with the CPO, we can use its taxonomic
structure for enrichment analyses. Similarly, once more cellular
phenotypes are being annotated in various model organism databases, we
can utilize CPO for integrative phenotype analyses across
species. However, at the moment, no comprehensive ontology of
physiology phenotypes is available, partly due to the complexity of
the domain and partly due to limitations in available ontology
frameworks. We see our main contribution in the development of a
theory that allows to capture physiology phenotypes; we applied this
theory to construct (largely automatically) the CPO, which is now
being used to annotate cellular phenotypes in the systems microscopy
domain. Based on our contributions, it is now possible to annotate
and eventually integrate (through the use of a controlled vocabulary
of physiology phenotypes, of which the CPO provides a first example)
data in the domain of physiology. At the same time, we maintain
compatibility with GO (on which a majority of physiology phenotypes
are based) and other PATO-based phenotype ontologies, thereby filling
a gap that has limited data integration in the physiology domain so
far.

Reviewer 2:
A theory of physiology related phenotypes and an implementation based
on GO, GO cross-products and PATO is presented. An outline of a theory
of phenotypes is introduced which participate-in a set of pathological
processes which can be qualified by a small set of PATO qualities. At
least, for a majority of heterogeneous phenotype descriptions this can
be an approach for unification. In future work a sound formal
characterization has and an evaluation of the practical value of the
approach has to be delivered.

Response:
We have now added a detailed formal presentation of our axioms as
supplementary material. An evaluation of the practical utility depends
on data that is being annotated with physiology phenotypes. Currently,
no comprehensive ontology of physiology phenotypes is available. Our
work brings the immediate benefit that the availability of CPO enables
annotation of cellular phenotypes in the systems microscropy
community, and our method can be applied to physiology phenotypes in
other domains, in particular computational modelling.

Reviewer 2:
Minor revision (page 2 line 27): You introduce a Heart beating example
to illustrate the challenges in modeling processes and their
regulation. [...] The Heart example is illustrative but limited with
respect to your ontology. There, both the the function of the single
entity can be impaired and the regulation of multiple occurences of
processes (where the example ends due to the single heartedness of
many animals).

Response: 
In response to this comment and reviewer 1's comment, we have removed
the Heart beating example from the introduction and focus on the
detailed presentation of the processes involved in the Results section
of our manuscript.

Reviewer 2:
Minor revision (page 3 second column line 33) see as above on the
phases of a single Heart action. The heart has no phase of
“inactivity” (that would be death), only a number of phases of which
the filling/ ejection phases are the least granular.

Response: We have changed our manuscript, according to the reviewer's
suggestion.

Reviewer 2:
Minor revision (page 5 Table 1): make it clear that you refer to the
product of row and column quantitites in the cells. I find it
difficult to understand this easy relation from your legend. Do you
really need this table?

Response:
We have removed the table from our manuscript, and discuss the
content of the table now in the main text.

Reviewer 2:
Class names should be consistently written italic and uppercase
throughout the manuscript (e.g. Heart beating and Rate of heart
beating)

Response:
We have made this change throughout our manuscript.

### 2012-04-04 trunk/paper/drinking-bioinformatics.tex

\bibliographystyle{plain}

\bibliography{/home/leechuck/Documents/papers/bibtex/lc}

>>>>

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{Ashburner2000short}
Michael Ashburner et~al.
\newblock Gene ontology: tool for the unification of biology.
\newblock {\em Nature Genetics}, 25(1), May 2000.

\bibitem{Bada2007a}
M.~Bada and L.~Hunter.
\newblock Enrichment of obo ontologies.
\newblock {\em Journal of Biomedical Informatics}, 40(3):300--315, June 2007.

\bibitem{Bada2004}
Michael Bada, Robert Stevens, Carole Goble, Yolanda Gil, Michael Ashburner,
  Judith~A. Blake, Michael~J. Cherry, Midori Harris, and Suzanna Lewis.
\newblock A short study on the success of the gene ontology.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web}, 1(2):235--240, February 2004.

\bibitem{Bard2005}
Jonathan Bard, Seung~Y. Rhee, and Michael Ashburner.
\newblock An ontology for cell types.
\newblock {\em Genome Biology}, 6(2), 2005.

\bibitem{Berners-Lee2001}
T.~Berners-Lee, J.~Hendler, O.~Lassila, et~al.
\newblock {The Semantic Web}.
\newblock {\em Scientific American}, 284(5):28--37, 2001.

\bibitem{Conrad2011}
Christian Conrad, Annelie W\"{u}nsche, Tze Heng~H. Tan, Jutta Bulkescher, Frank
  Sieckmann, Fatima Verissimo, Arthur Edelstein, Thomas Walter, Urban Liebel,
  Rainer Pepperkok, and Jan Ellenberg.
\newblock Micropilot: automation of fluorescence microscopy-based imaging for
  systems biology.
\newblock {\em Nature methods}, 8(3):246--249, March 2011.

\bibitem{opb}
Daniel~L. Cook, Fred~L. Bookstein, and John~H. Gennari.
\newblock Physical properties of biological entities: An introduction to the
  ontology of physics for biology.
\newblock {\em PLoS ONE}, 6(12):e28708, 12 2011.

\bibitem{Eilbeck2005}
K.~Eilbeck, S.~E. Lewis, C.~J. Mungall, M.~Yandell, L.~Stein, R.~Durbin, and
  M.~Ashburner.
\newblock The sequence ontology: A tool for the unification of genome
  annotations.
\newblock {\em Genome Biology}, 6(R55), 2005.

\bibitem{aposhort}
Stacia~R. Engel et~al.
\newblock Saccharomyces genome database provides mutant phenotype data.
\newblock {\em Nucleic acids research}, 38(Database issue), January 2010.

\bibitem{Fuchs2010}
Florian Fuchs, Gregoire Pau, Dominique Kranz, Oleg Sklyar, Christoph Budjan,
  Sandra Steinbrink, Thomas Horn, Angelika Pedal, Wolfgang Huber, and Michael
  Boutros.
\newblock Clustering phenotype populations by genome-wide {RNAi} and
  multiparametric imaging.
\newblock {\em Molecular Systems Biology}, 6, June 2010.

\bibitem{go2010}
{Gene Ontology Consortium}.
\newblock The gene ontology in 2010: extensions and refinements.
\newblock {\em Nucleic acids research}, 38(Database issue):D331--335, January
  2010.

\bibitem{Gkoutos2005}
Georgios~V. Gkoutos, Eain~C. Green, Ann-Marie~M. Mallon, John~M. Hancock, and
  Duncan Davidson.
\newblock {Using ontologies to describe mouse phenotypes.}
\newblock {\em Genome biology}, 6(1), 2005.

\bibitem{obml2011h1}
Georgios~V. Gkoutos and Robert Hoehndorf.
\newblock Ontology-based cross-species integration and analysis of
  saccharomyces cerevisiae phenotypes.
\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
  and Life sciences (OBML)}, October 2011.

\bibitem{Gkoutos2009b}
Georgios~V. Gkoutos, Chris Mungall, Sandra Dolken, Michael Ashburner, Suzanna
  Lewis, John Hancock, Paul Schofield, Sebastian Kohler, and Peter~N. Robinson.
\newblock Entity/quality-based logical definitions for the human skeletal
  phenome using {PATO}.
\newblock {\em Annual International Conference of the IEEE Engineering in
  Medicine and Biology Society.}, 1:7069--7072, 2009.

\bibitem{goble}
C.~Goble and R.~Stevens.
\newblock State of the nation in data integration for bioinformatics.
\newblock {\em Journal of Biomedical Informatics}, 41(5):687--693, 10 2008.

\bibitem{Grau2008}
B.~Grau, I.~Horrocks, B.~Motik, B.~Parsia, P.~Patelschneider, and U.~Sattler.
\newblock {OWL} 2: The next step for {OWL}.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web}, 6(4):309--322, November 2008.

\bibitem{Gruber1995}
Thomas~R. Gruber.
\newblock Toward principles for the design of ontologies used for knowledge
  sharing.
\newblock {\em International Journal of Human-Computer Studies}, 43(5-6), 1995.

\bibitem{Guarino1998}
Nicola Guarino.
\newblock Formal ontology and information systems.
\newblock In Nicola Guarino, editor, {\em Proceedings of the 1st International
  Conference on Formal Ontologies in Information Systems}, pages 3--15. IOS
  Press, 1998.

\bibitem{Herre2006}
Heinrich Herre, Barbara Heller, Patryk Burek, Robert Hoehndorf, Frank Loebe,
  and Hannes Michalek.
\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
  integrating objects and processes [{V}ersion 1.0].
\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany,
  2006.

\bibitem{Hoehndorf2010patterns}
Robert Hoehndorf, Anika Oellrich, Michel Dumontier, Janet Kelso, Dietrich
  Rebholz-Schuhmann, and Heinrich Herre.
\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
\newblock {\em BMC Bioinformatics}, 11(1):441+, 2010.

\bibitem{Hoehndorf2010phene}
Robert Hoehndorf, Anika Oellrich, and Dietrich Rebholz-Schuhmann.
\newblock Interoperability between phenotype and anatomy ontologies.
\newblock {\em Bioinformatics}, 26(24):3112 -- 3118, 10 2010.

\bibitem{Horridge2007}
Matthew Horridge, Sean Bechhofer, and Olaf Noppens.
\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
  OWL Experiences and Directions}, 2007.

\bibitem{Horrocks2007}
Ian Horrocks.
\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
  {W}eb {O}ntology {L}anguage.
\newblock Technical report, University of Manchester, March 2007.
\newblock \url{http://www.cs.man.ac.uk/~horrocks/obo/}.

\bibitem{Kazakov2011}
Yevgeny Kazakov, Markus Kr{\"o}tzsch, and Franti\v{s}ek Siman\v{c}\'{i}k.
\newblock Unchain my $\mathcal{EL}$ reasoner.
\newblock In {\em Proceedings of the 23rd International Workshop on Description
  Logics (DL'10)}, CEUR Workshop Proceedings. CEUR-WS.org, 2011.

\bibitem{Lock2010}
John~G Lock and Staffan Strömblad.
\newblock Systems microscopy: an emerging strategy for the life sciences.
\newblock {\em Experimental Cell Research}, 316(8):1438--1444, 2010.

\bibitem{owlprofiles}
Boris Motik, Bernardo~Cuenca Grau, Ian Horrocks, Zhe Wu, Achille Fokoue, and
  Carsten Lutz.
\newblock Owl 2 web ontology language: Profiles.
\newblock Recommendation, World Wide Web Consortium (W3C), 2009.

\bibitem{Mungall2010}
Christopher Mungall, Georgios Gkoutos, Cynthia Smith, Melissa Haendel, Suzanna
  Lewis, and Michael Ashburner.
\newblock Integrating phenotype ontologies across multiple species.
\newblock {\em Genome Biology}, 11(1):R2+, 2010.

\bibitem{Mungall2010go}
Christopher~J. Mungall, Michael Bada, Tanya~Z. Berardini, Jennifer Deegan,
  Amelia Ireland, Midori~A. Harris, David~P. Hill, and Jane Lomax.
\newblock Cross-product extensions of the gene ontology.
\newblock {\em Journal of biomedical informatics}, February 2010.
\newblock in press.

\bibitem{Neumann2010}
Beate Neumann, Thomas Walter, Jean-Karim Hériché, Jutta Bulkescher, Holger
  Erfle, Christian Conrad, Phill Rogers, Ina Poser, Michael Held, Urban Liebel,
  and et~al.
\newblock Phenotypic profiling of the human genome by time-lapse microscopy
  reveals cell division genes.
\newblock {\em Nature}, 464(7289):721--727, 2010.

\bibitem{Ogren2004}
P.~V. Ogren, K.~B. Cohen, G.~K. Acquaah-Mensah, J.~Eberlein, and L.~Hunter.
\newblock The compositional structure of gene ontology terms.
\newblock {\em Pac Symp Biocomput}, pages 214--225, 2004.

\bibitem{Gruninger2010}
Atalay \"{O}zg\"{o}vde and Michael Gr\"{u}ninger.
\newblock Foundational process relations in bio-ontologies.
\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
  Information Systems}, pages 243--256, Amsterdam, The Netherlands, The
  Netherlands, 2010. IOS Press.

\bibitem{Robinson2008}
P.~N. Robinson, S.~Koehler, S.~Bauer, D.~Seelow, D.~Horn, and S.~Mundlos.
\newblock The human phenotype ontology: a tool for annotating and analyzing
  human hereditary disease.
\newblock {\em American journal of human genetics}, 83(5):610--615, 2008.

\bibitem{wpo}
Gary Schindelman, Jolene Fernandes, Carol Bastiani, Karen Yook, and Paul
  Sternberg.
\newblock Worm phenotype ontology: integrating phenotype data within and beyond
  the c. elegans community.
\newblock {\em BMC Bioinformatics}, 12(1):32, 2011.

\bibitem{Schmitz2010short}
Michael H.~A. Schmitz et~al.
\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
  importin-$\beta$1 as key mitotic exit regulators in human cells.
\newblock {\em Nature Cell Biology}, 12:886--893, 2010.

\bibitem{Schofield2011}
Paul~N. Schofield, John~P. Sundberg, Robert Hoehndorf, and Georgios~V. Gkoutos.
\newblock New approaches to the representation and analysis of phenotype
  knowledge in human diseases and their animal models.
\newblock {\em Briefings in Functional Genomics}, 10(5):258--265, 2011.

\bibitem{Smith2007short}
Barry Smith et~al.
\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
  biomedical data integration.
\newblock {\em Nat Biotech}, 25(11):1251--1255, 2007.

\bibitem{Smith2004}
Cynthia~L. Smith, Carroll-Ann~W. Goldsmith, and Janan~T. Eppig.
\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
  and comparing phenotypic information.
\newblock {\em Genome Biology}, 6(1):R7, 2004.

\end{thebibliography}

% \bibliography{/home/leechuck/Documents/papers/bibtex/lc}
